---
title: "DM_training"
output: html_document
date: "2023-10-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### As people are entering, direct them to download the park_visits.csv from DM Training page and save it to their computer. Then open RStudio. 

## Class logistics ---
  - TRAINERS INTRODUCE themselves very briefly
  
  - We are assuming you HAVE NEVER OR HAVE BARELY R before
  
  - If you get lost, have questions, etc., RAISE YOUR HAND AND A FLOATING TRAINER WILL COME TO HELP YOU
  
  - Everything we go over in class will be provided in DETAILED CLASS NOTES that will be available for you to download from the DM Training website within the next few days. 
  
  - If you want to continue learning R after this class I strongly suggest you work through the R training materials from our SCIENTISTS' TRAINING earlier this year [** SHOW]. We did 20 hours of Introductory R for that training, and everything is available at this website. I'll include the URL on the class notes.
  
  - Today, our preference is for you to CODE ALONG WITH US during the next hour plus. I'll be typing code up here at a pace that will allow you to type along on your computer so you can make sure you're getting the expected outputs from R. 
  
  - We are available to help you with any of this. If you get hopelessly lost with the class materials today, we can SET UP A TEAMS MEETING after the DM training is over. Just reach out to us. The TRAINER EMAILS WILL BE IN THE CLASS NOTES.

## Goals for Today ---
With this introduction to R we hope to give you a sense of how R works, so you have a STARTING POINT FROM WHICH YOU CAN CONTINUE LEARNING R--if you want--from the SCIENTISTS' TRAINING WEBSITE or the many FREE R TUTORIALS that are available on the Internet. 

Try not to get frustrated if you don't understand everything today or if you can't get some code to work. R CAN BE A LITTLE DIFFICULT TO LEARN, but if you KEEP AT IT you will find that it is an incredibly POWERFUL PROGRAMMING LANGUAGE for cleaning and manipulating data, analyzing data, and producing really nice data visualizations.

Note that there are MANY DIFFERENT WAYS in R. If we want to SUBSET A LARGE DATA SET to just get the data for a particular park, one person might use R's subset() function to perform that task while another may use the filter() function to achieve the very same thing. Use the functions you are most comfortable with to do what you need with the data.

FROM THE PRE-CLASS VIDEOS you learned how to:
  - navigate the RStudio interface
  - install and load R packages

BY THE END OF THIS (VERY SHORT) CLASS you should be able to:
  - create and save R projects and R scripts
  - import a .csv file, perform basic data cleaning and manipulations, and export the CLEANED data as a .csv file
  - create a plot in R and export it as an image file

## Create an RStudio project ---
** FILE > NEW PROJECT > NEW DIRECTORY (and browse to location) -- project extension is .Rproj. You can also use the New File icon to create a new RStudio project.

WHY USE AN RSTUDIO PROJECT? We can create and run an R script without associating the script with an RStudio project. But with an RStudio project...

  - All the files that you generate from your R script should be saved in the project folder or in a project subfolder. This is a good way to ORGANIZE ALL YOUR R SCRIPTS AND OUTPUTS for a project. R will use the location of the RStudio project as your DEFAULT WORKING DIRECTORY, whenever you're working in the project.

```{r}
getwd()
```

TO OPEN THIS PROJECT the next time you want to work in it, just double-click the .Rproj file and it will automatically open your project in RStudio and you'll see all your project-related files as well.

IF YOU WANT TO SHARE THIS RSTUDIO PROJECT WITH A COLLEAGUE, you can just zip the entire folder and send it.

## Coding in R ---
Let's start with some CODING IN THE RSTUDIO CONSOLE

As you learned in the pre-class video, when you type CODE IN THE CONSOLE and then press enter, the code WILL EXECUTE IMMEDIATELY. 

But before we start coding, two points:
  - R IS CASE SENSITIVE
  - It's a good idea to adopt a consistent, CLEAN CODING STYLE for all your projects. Google has an R Style Guide that I like. It's a modification of the Tidyverse R Style Guide [** GOOGLE TO SHOW]

Let's do some math in the R console

```{r}
2 + 3 # JUST PRESS ENTER. The SPACING is just for ease of reading.

x <- 4 # If you want to ASSIGN A VALUE TO A VARIABLE in R, standard practice is to use this arrow ASSIGNMENT OPERATOR rather than the equal sign though both will work in most situations. 
```

Notice that when you ran that line of code it DIDN'T PRINT ANY OUTPUT in the console, it only assigned the number to the variable x and saved it in it memory. 

You can see that R now has this variable in its GLOBAL ENVIRONMENT. 

For the remainder of this R session, R will recognize x as equal to 4 unless you overwrite it by assigning a different value to x or if you remove that variable from memory with the rm() function. 

```{r}
x * 5 # You can do math. NOT ASSIGNING THE OUTPUT TO A VARIABLE

# Let's calculate the mean of 4 numbers
mean_val <- (2 + 4 + 5 + 8)/4' # R respects ORDER OF OPERATIONS, so you need to enclose the additions in parentheses to do that first before dividing by 4. 
```

## Working with Vectors ---
So far we have only assigned values to variables. In R, the VARIABLE IS CONSIDERED A VECTOR OF LENGTH 1. A VECTOR IS THE MOST BASIC DATA STRUCTURE in R. You can think of a vector as a SEQUENCE OF VALUES THAT HAVE THE SAME DATA CLASS, e.g., a collection of numbers or a collection of character strings.

Create a vector of 4 numbers:

```{r}
x <- c(2, 4, 5, 8) # You'll see in the GLOBAL ENVIRONMENT that R now recognizes the variable 'x' as a vector of four items with a NUMERIC DATA CLASS. Notice we had to wrap the four numbers in--c-parentheses--to create the vector. c() is an R FUNCTION FOR COMBINING VALUES to create a vector. 

x <- 2, 4, 5, 8 # ERROR MESSAGE
```

When you have a vector of many numbers, you can perform MATH CALCULATIONS on them the same way you did when x was a single number. 
```{r}
x * 3 # This can be useful if, for example, you need to convert a vector of fish lengths in inches to fish lengths in cm.
```  

Let's learn some R FUNCTIONS commonly used with vectors: 
  
```{r}
class(x) # to show the data class. Other data classes in R include character data class for string text and logical for TRUE/FALSE values.

# What if I want to see HOW LONG this vector is? I can look in the global environment, but R also has a base R function for this task. 

# ** GOOGLE 'r how do i find the number of values in a vector' 
length(x)
```

CHALLENGE <<<<<<
Figure out how to find the minimum value, mean value, and standard deviation of your vector of numbers...just like we did to find the length() function... figure out the R functions to use, run those functions on your vector and see what the output is. 

```{r}
  
min(x)
mean(x) # applying the mean function to your vector of numbers is easier than calculating the mean by yourself.
sd(x) 
```

## Functions---

LET'S TALK ABOUT FUNCTIONS IN MORE DETAIL NOW b/c functions are super important in R...

```{r}
?mean # R-help will open the documentation that describes this function
```
  
In R, question-mark-function-name is an easy way to find the HELP FILES for an R function. We see this is a function in BASE R. That means it's part of the collection of functions automatically provided with an R installation, and it automatically loads up when you open RStudio. 

If the function is part of an R PACKAGE that you installed, you would see the package name in the squigglies. 

The help file explains what this function does. R help files are generally DIFFICULT TO UNDERSTAND, so don't worry if this looks unintelligible to you. I often do Google searches to get with with R, instead of using R help files.

### What is a function in R? ---
 
You can think of a function as a SEQUENCE OF R COMMANDS that typically 
  - accepts INPUT values, 
  - runs a sequence of pre-written R COMMANDS on those input values,      - then produces some OUTPUT

R PACKAGES ARE COMPILATIONS OF FUNCTION and example datasets that someone has pre-written for you. 
  
WE USE FUNCTIONS BY TYPING THEIR NAME AND A SET OF PARENTHESES.

ARGUMENTS, WHICH ARE THE INFORMATION YOU GIVE THE FUNCTION ABOUT YOUR DATA, GO INSIDE THE PARENTHESES.

IT'S BAD CODING PRACTICE TO USE FUNCTION NAMES AS VARIABLE NAMES, SO TRY TO AVOID IT.

You can CREATE YOUR OWN FUNCTION IN R and then call your custom function every time you need to perform that task in R. We don't have time to go over it today, but it's explained in detail in the Scientists Training website.

### Default argument values ---
Notice that in the Usage explanation for mean, there are THREE ARGUMENT FUNCTIONS shown and two of them have associated values. These two--TRIM and NA.RM--are ARGUMENTS THAT HAVE DEFAULT VALUES. That means if you don't enter any value for these arguments, the function will run using the default values. The ONLY REQUIRED ARGUMENT for this function is x, and x is the R object, for example the vector of numbers.
  
  The function ARGUMENTS are shown in A PARTICULAR ORDER. If you call a function and enter arguments in a different order than shown in the arguments list, then you need to provide the argument name as well. Let's see an example.
  
```{r}
# Replace the 5 in your x-vector with NA, which as a reserved word in R, that stands for missing value. Reserved words cannot be used as variable names in R. 

x <- c(2, 4, NA, 8)
mean(x) # R returns NA as the mean of the vector. 

# Check the help file...
mean(x, na.rm = TRUE) # In this case, I provided the argument name followed by the argument value: na.rm = TRUE. That's because this is the third argument in the function, but I'm entering it as the second argument. See what happens if I don't include the argument name.

mean(x, TRUE)
```
  

Let's look in more detail ANOTHER EXAMPLE--A PLOTTING FUNCTION. Use this as a way to INTRODUCE YOU TO PLOTS AND ALSO TO ANOTHER DATA STRUCTURE, THE DATA FRAME.
  
```{r}

?boxplot # an R function to create boxplots from your data. The graphics package is default loaded with R, as a base package.
```

Let's SCROLL DOWN TO THE BOTTOM of the help page and SEE THE EXAMPLES provided for this function. Often I find this an easy way to understand how to use a function. 

Help examples often use EXAMPLE DATASETS that are provided with R, or example data sets that are provided with the package itself. In this first example, we see that the data used it called InsectSprays.

```{r}

?InsectSprays 
data("InsectSprays") # Expand in global environment

# This is a data frame, which is the next data structure we will learn so I'll just touch on it briefly now.

# A data frame is a TABULAR DATA STRUCTURE.

# EACH COLUMN IS A VECTOR WITH AN ASSOCIATED DATA CLASS. Typically, each column of a data frame is an observation--this is what we would call a 'tidy' data frame and it's the preferred way to organize your data for R.

# We haven't seen the FACTOR DATA CLASS yet--it means that the spray column is a categorical variable with a limited number of levels. In this case, 6 levels. These are the only values that can be entered in this column.

# CLICK THE EXCEL ICON AND SEE HOW TO SORT AND FILTER

# LET'S GO BACK TO THE BOXPLOT HELP PAGE AND SEE HOW THE EXAMPLE FEEDS THIS DATA FRAME INTO THE FUNCTION TO GENERATE A BOXPLOT.

?boxplot

# HIGHLIGHT THE FIRST EXAMPLE AND PASTE IT IN YOUR CONSOLE, THEN PRESS ENTER. 
boxplot(count ~ spray, data = InsectSprays, col = "lightgray")

```

CHALLENGE <<<<<<<<<<
Use the function help page to figure out how to make these changes to the boxplot (and try them out):
- convert the plot from vertical boxplots to horizontal
- the boxplots should be orange
- and then use the examples to figure out how to add a title to your plot
```{r}
boxplot(count ~ spray, data = InsectSprays, col = "orange", horizontal = TRUE, main = "Effectiveness of Insect Sprays") # notice the function AUTOFILL
```

SAVE THE PLOT:
(RESIZE PLOTS PANE) EXPORT > SAVE AS IMAGE

## R script ---
We are going to switch now from typing R commands in the console to typing commands in an R Script.

An R script is just a TEXT FILE WHERE YOU CAN TYPE R COMMANDS AND COMMENT YOUR CODE. You can RUN COMMANDS FROM AN R SCRIPT, and you can also save the R script so you can re-run it later without having to type everything over again.

Let's create an R script that will be associated with this project. 
** FILE > NEW FILE > R SCRIPT. OR.... use the SHORTCUT ICON

  - In R scripts, use the HASHTAG TO IDENTIFY A COMMENT. When executing code, R will ignore anything that starts with a hashtag. 

  - I always start my R scripts with information about the PURPOSE OF THE SCRIPT
  
  - At the beginning of a script I will also load any packages I will need. 
  ** Type 'library(tidyverse)' in script. 
  
  EXPLAIN HOW TO EXECUTE CODE LINES by putting your cursor on that line and press Enter, or highlight the code and press Enter (or push 'Run'--explain that 'SOURCE' will run the entire script). 
  
  Notice that WHEN YOU LOAD A PACKAGE, you will see a checkmark to the left of it in the RStudio Packages tab.
  
  - I also like to organize my script with COLLAPSIBLE SECTIONS for easier navigation. [** SHOW]
  
  - Now use the 'SAVE' icon to save the script in your project working directory. You can also use FILE > SAVE AS to save the script somewhere else on your computer.
  
## PARK VISITS ---
We're going to IMPORT THE PARK_VISITS .csv file and examine the data in R.

When you import a .csv file into R, it is by default IMPORTED AS A DATA FRAME

BRIEFLY EXPLAIN THE PARK_VISITS DATA SET

** (Files tab) click the file you want, then IMPORT DATASET (default will be as a data frame) Gives you a preview of your data, its guesses on the data type for each column, default data frame name it will import as, and the actual R code used to import the file with the selections shown here. 

For our demo we will just CHANGE THE NAME TO 'DAT' and then import with the rest of the default settings. R will import the data, save it as a data frame named 'dat', and in the console it will show you the code. 

The import function also automatically calls the VIEW() function to disply the data as a tab in the RStudio source pane.

In the GLOBAL ENVIRONMENT you'll see the data frame you just imported. 

  - Notice that the YEAR COLUMN has been imported as a CHARACTER DATA CLASS--this suggests there are some entries in this column that are not numeric

 - Because each column of a data frame is considered a vector, you can also use the class() function to find the data type for a column of a data frame. 

Let's run some R functions commonly used with data frames: 

```{r}
summary(dat) # show SUMMARY INFORMATION FOR EACH COLUMN. The summary information is not very interesting for the character data classes, but the visitors column is a numeric data class and the summary() function shows us the minimum, maximum, and some other useful information for this column of data.

table(dat$region) # The dollar sign followed by column name is one way to access a single column of a data frame.

unique(dat$year) # see something funny in there
```

## Clean/manipulate the parks data ---
Now we are going to start using FUNCTIONS FROM THE TIDYVERSE PACKAGE to clean and manipulate the data for plotting.

The 'tidyverse' package is really a COLLECTION OF MANY R PACKAGES DESIGNED TO WORK TOGETHER TO SIMPLIFY DATA SCIENCE TASKS. 

Tidyverse functions use a coding style called 'PIPING' TO CHAIN TOGETHER A SEQUENCE OF R COMMANDS in a way that makes the R code easier to read.

Let's say we want to modify the parks data by first removing all data records where the year is entered as "Total" and then we want to convert the year column from character to integer data class.

This is how we would do that in tidyverse style.
```{r}
# Start with filtering the data to only include records where 'year' is NOT equal to 'Total'
subdat <- dat %>% dplyr::filter(year != "Total") # PIPING OPERATOR

unique(subdat$year)

subdat <- subdat %>% dplyr::mutate(year = as.integer(year))
unique(subdat$year) # no quotes! Check ENVIRONMENT PANE

# NOW PIPE IT ALL TOGETHER
rm(subdat)

subdat <- dat %>% 
  dplyr::filter(year != "Total") %>% # ...AND THEN...
  dplyr::mutate(year = as.integer(year))

```

R TIDYVERSE CHEATSHEET

SCIENTISTS' TRAINING PAGE

The functions we used above are from the DPLYR PACKAGE, which is one of the packages loaded with TIDYVERSE.

Let's look at the documentation for this package to see what this package is all about. ** PACKAGES > SEARCH 'dplyr' > USER GUIDE

WE'RE GOING TO PIPE A FEW MORE DPLYR FUNCTIONS TO FINISH CLEANING THE PARKS DATA SET.

OUR GOAL IS TO EXTRACT A SUBSET OF THE DATA AND MANIPULATE IT SO WE CAN CREATE THIS PLOT IN R [** SHOW and explain plot]. This plot shows the number of park visitors per year in national parks from the Midwest region. 

```{r}
rm(subdat)

subdat <- dat %>%
  dplyr::filter(region == "MW" & year != "Total")
```

NEXT WE WILL...
- Reduce the data to include only these columns of data: year, unit_code, visitors
- Convert the year column to an integer data type (we already demo'ed)
- Create a new column called 'visitors_1000' that is calculated by dividing visitors by 10000
```{r}
subdat <- dat %>%
  dplyr::filter(region == "MW" & year != "Total") %>% 
  dplyr::select(year, unit_code, unit_name, visitors) %>%
  dplyr::mutate(
    year = as.integer(year),
    visitors_1000 = visitors/1000) 
```

CHALLENGE >>>>> ADD ONE MORE PIPE TO SORT THE DATA BY UNIT_CODE, THEN BY YEAR. FIGURE OUT THE DPLYR FUNCTION FOR SORTING DATA.
```{r}
subdat <- dat %>%
  dplyr::filter(region == "MW" & year != "Total") %>% 
  dplyr::select(year, unit_code, unit_name, visitors) %>%
  dplyr::mutate(
    year = as.integer(year),
    visitors_1000 = visitors/1000) %>%
  dplyr::arrange(unit_code, year) 
```

SAVE YOUR CLEANED DATA AS A .CSV FILE
```{r}
write_csv(subdat, "Midwest_park_visits.csv")
```

BONUS >>>>
```{r}
total_by_park <- subdat %>%
  dplyr::group_by(unit_code) %>%
  summarize(tot_visitors_1000 = sum(visitors_1000))
```

BONUS CHALLENGE>>>
```{r}
barplot(tot_visitors_1000 ~ unit_code, data = total_by_park)
```

## CREATE A LINE GRAPH, USING GGPLOT
# ?geom_line examples. Find one that colors the lines by a variable.
```{r}
?geom_line # You could have googled to figure this out

# RUN THE FIRST TWO EXAMPLES

# Notice the use of PLUS, not PIPE
```

CHALLENGE <<<<<
NOW CREATE A LINE GRAPH WITH YOUR PARKS DATA (DON'T WORRY ABOUT ADDING TITLES, ETC. FOR NOW)
X-axis = year, Y-axis = visitors_1000, color by unit_code
```{r}
# THE EXAMPLE
ggplot(economics_long, aes(date, value01, colour = variable)) +
  geom_line() # This was the example. Just replace it with columns from your data.

# YOUR PLOT
ggplot(subdat, aes(year, visitors_1000, colour = unit_code)) +
  geom_line() 
  
```

Pretty it up a bit with nicer title and axis labels
```{r}
visitors_plot <- ggplot(subdat, aes(x = year, y = visitors_1000, colour = unit_code)) + 
  geom_line() +
  labs(title = "Annual number of visitors to National Parks in Midwest Region", x = "Year", y = "Visitors (1000's)") +
  theme_bw()
```

FOR FUN, MAKE IT INTERACTIVE!!
```{r}
plotly::ggplotly(visitors_plot)

# Plotly perks
# > hover info.
# > select unit_codes to show and hide
# > download plot
# > compare data on hover

```